Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-08-16 paper/paperv3.qmd
Last few changes to supplemental plots

### Phenology

@@

mean_yb_phen <- pheno_raw %>% 
  filter(age == 4, phase == "breeddur") 
mean_ob_phen <- pheno_raw %>% 
  filter(age == 11, phase == "breeddur")
mean_ym_phen <- pheno_raw %>% 
  filter(age == 4, phase == "moltdur")
mean_om_phen <- pheno_raw %>% 
  filter(age == 11, phase == "moltdur") 
```

Overall, the breeding haul out was `r mean_yb_phen$dur_days` (`r mean_yb_phen$dur_days_lwr` - `r mean_yb_phen$dur_days_upr`) days for 4 year olds and `r mean_ob_phen$dur_days` (`r mean_ob_phen$dur_days_lwr` - `r mean_ob_phen$dur_days_upr`) days for 11 year old animals. The molting haul out was `r mean_ym_phen$dur_days` (`r mean_ym_phen$dur_days_lwr` - `r mean_ym_phen$dur_days_upr`) days for 4 year olds and `r mean_om_phen$dur_days` (`r mean_om_phen$dur_days_lwr` - `r mean_om_phen$dur_days_upr`) days for animals 11 years old and older.

@@

ggplot(selective_coef, aes(model, mean, color = hypothesis)) + 
  geom_pointrange(aes(ymin = lwr, ymax = upr), position = position_dodge(width = 0.2)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  expand_limits(y = 0) + 
  labs(y = "Coefficient", x = "Model") + 
  theme(legend.position = "none")

@@

## Observation frequency

@@

```{r}
#| label: missing-seals

## Are we missing animals more or less with age?
missing_seals <- sealdat %>% 
  filter(age <= 20) %>% 
  # Fill in non-resighted years
  group_by(animalID) %>% 
  arrange(year) %>% 
  reframe(tibble(year2 = (yearborn[1] + 4):max(year), # from age 4 to last obs
                 unobserved = as.integer(!year2 %in% year),
                 age = year2 - yearborn[1],
                 age10 = (age - age_senesce) / 10)) %>% 
  ungroup() %>% 
  rename(year = year2) %>% 
  filter(year >= 1988) %>% 
  mutate(age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old")),
         year_fct = factor(year)) %>% 
  left_join(select(sealdat, animalID, year, observed_int),
            by = c("animalID", "year"))

# Summary: mean and 95% CI 
missing_summ <- missing_seals %>% 
  group_by(age, age_cat) %>% 
  summarize(n = n(),
            n_unobserved = sum(unobserved),
            unobserved_mean = n_unobserved / n,
            unobserved_lwr = binom.test(n_unobserved, n)$conf.int[1],
            unobserved_upr = binom.test(n_unobserved, n)$conf.int[2],
            .groups = "drop")

missing_mod <- glmer(
  unobserved ~ age10 + (1 | year_fct) + (1 | animalID), 
  missing_seals, 
  family = binomial
)

summary(missing_mod)

@@

First, we tested whether probability of detection was related to age. We found the proportion of unobserved animals decreased with age (i.e., older seals are more likely to be observed). For this analysis, we excluded the last year each animal was observed. Then, for all years between animals' age 4 year to the year prior to final observation, we determined whether each animal was observed. We estimated the effect of age on probability of an animal being unobserved using a GLMM (binomial response, logit link) with random effects for year and animal ([@fig-unobserved]). The coefficient for age was -0.2593 (95% CI: -0.4896 - -0.0290, *p* = 0.027). The change in mean unobserved rate from age 4 to 20 was 20.0% to 14.2%.

```{r}
#| label: fig-unobserved
#| fig-cap: The probability of an animal being unobserved decreased with age. Points and error bars represent the proportion of unobserved animals with 95% CI. The blue line and shaded area are the mean and 95% CI probability of an animal being unobserved (GLMM, with random effect of year weighted by number of seals observed each year). The dotted line is the mean response of the unweighted predictions. Numbers at the top indicate number of seals in each age category. See text for model details.

# Prediction grid
missing_grid <- expand_grid(age = 4:20, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10)

# Missing seal predictions (weighted model)
missing_pred <- missing_grid %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             re.form = ~ (1 | year_fct)),
         pred_pop = predict(missing_mod, 
                            newdata = .,
                            type = "response",
                            re.form = NA),
         se = suppressWarnings(predict(missing_mod, 
                                       newdata = .,
                                       se.fit = TRUE,
                                       re.form = NA)$se.fit),
         conf.low = predicted - 1.96 * se,
         conf.high = predicted + 1.96 * se) %>% 
  left_join(count(missing_seals, year_fct), by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(across(c(predicted, pred_pop, conf.low, conf.high), mean)) %>% 
  mutate(across(c(predicted, conf.low, conf.high), 
                family(missing_mod)$linkinv))

# Yearly predictions
missing_pred_years <- missing_grid %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             type = "response",
                             re.form = ~ (1 | year_fct)))

ggplot(missing_pred, aes(age)) +
  # Line per year
  geom_line(aes(y = predicted, group = year_fct), 
            missing_pred_years, 
            alpha = 0.2) +
  # Weighted model
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.3,
              fill = "cornflowerblue") +
  geom_line(aes(y = predicted),
            linewidth = 2,
            color = "cornflowerblue") +
  # Population-level model
  geom_line(aes(y = pred_pop), 
            linetype = "dotted",
            linewidth = 1.2,
            color = "cornflowerblue") +
  # Raw data
  geom_pointrange(aes(y = unobserved_mean, 
                      ymin = unobserved_lwr, 
                      ymax = unobserved_upr),
                  missing_summ) +
  # Sample size
  geom_text(aes(age, label = n), missing_summ, y = 0.8, vjust = 1.1) +
  scale_x_continuous("Age", breaks = seq(4, 20, by = 4)) +
  scale_y_continuous("Unobserved proportion", 
                     labels = scales::label_percent(accuracy = 1),
                     limits = c(0, NA)) +
  theme(legend.position = "none")
```

@@

## Population dynamics?

>>>>

### H4: Phenology

@@

pheno_linkinv <- family(pheno_model)$linkinv
expand_grid(
  age = c(4, 10),
  phase = c("moltdur", "breeddur")
) %>% 
  mutate(age_cat = factor(age >= age_senesce, 
                          labels = c("Young")),
         age10 = (age - age_senesce) / 10) %>% 
  mutate(dur_days_eta = predict(pheno_model, newdata = ., re.form = NA),
         dur_days_se = suppressWarnings(predict(pheno_model, 
                                                newdata = ., 
                                                re.form = NA, 
                                                se.fit = TRUE)$se.fit),
         dur_days = pheno_linkinv(dur_days_eta),
         dur_days_lwr = pheno_linkinv(dur_days_eta - 1.96 * dur_days_se),
         dur_days_upr = pheno_linkinv(dur_days_eta + 1.96 * dur_days_se)) %>% 
  select(-age10, -dur_days_eta, -dur_days_se)

@@

ggplot(selective_coef, aes(hypothesis, mean, color = model)) + 
  geom_pointrange(aes(ymin = lwr, ymax = upr), position = position_dodge(width = 0.2)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  expand_limits(y = 0) + 
  labs(y = "Coefficient", x = "Model") + 
  theme(legend.title = element_blank())

@@

## Detection frequency

@@

```{r}
#| label: seal-detections

# Are seal detections changing with age?
detection_seals <- sealdat %>% 
  filter(age <= 20) %>% 
  # Fill in non-resighted years
  group_by(animalID) %>% 
  # Keep only animals seen at least 2 years apart (i.e. if only seen in consecutive years remove them)
  mutate(year_range = max(year) - min(year)) %>% 
  filter(year_range > 2) %>% 
  select(-year_range) %>% 
  # Interpolate seen/not seen
  reframe(tibble(year2 = (yearborn[1] + 4):(max(year) - 1), # from age 4 to year prior last obs
                 observed = as.integer(year2 %in% year),
                 age = year2 - yearborn[1],
                 age10 = (age - age_senesce) / 10)) %>% 
  ungroup() %>% 
  rename(year = year2) %>% 
  filter(year >= 1988) %>% 
  mutate(age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old")),
         year_fct = factor(year)) %>% 
  left_join(select(sealdat, animalID, year, observed_int),
            by = c("animalID", "year"))

# Summary: mean and 95% CI 
detection_summ <- detection_seals %>% 
  group_by(age, age_cat) %>% 
  summarize(n = n(),
            n_observed = sum(observed),
            observed_mean = n_observed / n,
            observed_lwr = binom.test(n_observed, n)$conf.int[1],
            observed_upr = binom.test(n_observed, n)$conf.int[2],
            .groups = "drop")

detection_mod <- glmer(
  observed ~ age10:age_cat + (1 | year_fct) + (1 | animalID), 
  detection_seals, 
  family = binomial
)

summary(detection_mod)

@@

First, we tested whether detection probability was related to age. We found detection probalities decreased with age (i.e., older seals are less likely to be detected). For this analysis, we excluded the last year each animal was observed. Then, for all years between animals' age 4 year to the year prior to final observation, we determined whether each animal was detected. We estimated the effect of age on detection probability using a GLMM (binomial response, logit link) with an interaction between age and age category (young = age \< 11, old = age \> 11) and random effects for year and animal ([@fig-unobserved]). The coefficient for age was not significantly different than 0 for animals older than 11, but detection probability decreased significantly with age up to age 11. The change in detection probability from age 11 to 19 was 70.1% (61.7 - 77.3%) to 59.9% (44.4% - 73.7%).

```{r}
#| label: fig-unobserved
#| fig-cap: Detection probabilities decreased with age for young, but not old, animals. Points and error bars represent the proportion of detected animals with 95% CI. The green and pink lines and shaded areas are the mean and 95% CI probability of an animal being detected (GLMM with random effect of year, weighted by number of seals observed each year), color-coded by age category. The dotted line is the mean response of the unweighted predictions. Numbers at the top indicate number of seals of each age. See text for model details.

# Prediction grid
detection_grid <- expand_grid(age = 4:19, 
                              year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10,
         age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old")))

# Seal detection predictions (weighted model)
detection_pred <- detection_grid %>% 
  mutate(predicted = predict(detection_mod, 
                             newdata = .,
                             re.form = ~ (1 | year_fct)),
         pred_pop = predict(detection_mod, 
                            newdata = .,
                            type = "response",
                            re.form = NA),
         se = suppressWarnings(predict(detection_mod, 
                                       newdata = .,
                                       se.fit = TRUE,
                                       re.form = NA)$se.fit),
         conf.low = predicted - 1.96 * se,
         conf.high = predicted + 1.96 * se) %>% 
  left_join(count(detection_seals, year_fct), by = "year_fct") %>% 
  group_by(age, age_cat) %>% 
  summarize(across(c(predicted, pred_pop, conf.low, conf.high), mean), .groups = "drop") %>% 
  mutate(across(c(predicted, conf.low, conf.high), 
                family(detection_mod)$linkinv))

# Yearly predictions
detection_pred_years <- detection_grid %>% 
  mutate(predicted = predict(detection_mod, 
                             newdata = .,
                             type = "response",
                             re.form = ~ (1 | year_fct)))

ggplot(detection_pred, aes(age, color = age_cat, fill = age_cat)) +
  # Line per year
  geom_line(aes(y = predicted, group = interaction(age_cat, year_fct)), 
            detection_pred_years, 
            alpha = 0.2) +
  # Weighted model
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.3, 
              color = NA) +
  geom_line(aes(y = predicted, linetype = age_cat),
            linewidth = 1.5) +
  # Population-level model
  geom_line(aes(y = pred_pop), 
            linetype = "dotted",
            linewidth = 1.1) +
  # Raw data
  geom_pointrange(aes(y = observed_mean, 
                      ymin = observed_lwr, 
                      ymax = observed_upr),
                  detection_summ) +
  # Sample size
  geom_text(aes(age, label = n), detection_summ, y = 0, vjust = 0) +
  scale_x_continuous("Age", breaks = seq(4, 19, by = 4)) +
  scale_y_continuous("Detection rate", 
                     labels = scales::label_percent(accuracy = 1),
                     limits = c(0, NA)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  theme(legend.position = "none")
```

@@

## Population dynamics
